Quantifying the Relationship Between Genetic Diversity and Population Size Suggests Natural Selection Cannot Explain Lewontin's Paradox

### 2021-08-04 manuscript/elife/elife-manuscript.tex
minor title fix
edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Quantifying Lewontin's Paradox Suggests Natural Selection is Unlikely to
Explain the Narrow Range of Diversity Among Species}

@@

 Neutral theory predicts that genetic diversity increases with population size,
 yet observed levels of diversity across metazoans vary only two orders of
 magnitude while population sizes vary over several. This unexpectedly narrow
 range of diversity is known as Lewontin's Paradox of Variation (1974). While
 some have suggested selection constrains diversity, tests of this hypothesis
 seem to fall short. Here, I revisit Lewontin's Paradox to assess whether
 current models of linked selection are capable of reducing diversity to this
 extent. To quantify the discrepancy between pairwise diversity and census
 population sizes across species, I combine previously-published estimates of
 pairwise diversity from 172 metazoan taxa with estimates of census sizes.
 Using phylogenetic comparative methods, I show this relationship is
 significant accounting for phylogeny, but with high phylogenetic signal and
 evidence that some lineages experience shifts in the evolutionary rate of
 diversity deep in the past. Additionally, I find a negative relationship
 between recombination map length and census size, suggesting abundant species
 have less recombination and experience greater reductions in diversity due to
 linked selection. However, I show that even using strong selection parameter
 estimates, models of linked selection are unlikely to explain the observed
 relationship between diversity and census sizes across species.
\end{abstract}

@@

A longstanding mystery in evolutionary genetics is that the observed levels of
genetic variation across sexual species are confined to an unexpectedly narrow
range. Under neutral theory, the average number of nucleotide differences
between lineages (pairwise diversity, $\pi$) is determined by the balance of
new mutations and their loss by genetic drift
\citep{Kimura1964-ul,Malecot1948-zv,Wright1931-fl}. In particular, the expected
diversity at neutral sites in a panmictic population of $N_c$ diploids is
expected to be $\pi \approx 4N_c \mu$, where $\mu$ is the per basepair per
generation mutation rate. Given that metazoan germline mutation rates only
differ 10-fold ($10^{-8}$--$10^{-9}$, \cite{Kondrashov2010-fi,Lynch2010-ki}),
and census sizes vary over several orders of magnitude, one would expected
under neutral theory that heterozygosity should also vary over several orders
of magnitude. However, early allozyme surveys revealed that heterozygosity
levels across a wide range of species varied just an order of magnitude
\citep[][p. 208]{Lewontin1974-jb}; this is known as Lewontin's ``Paradox of
Variation". With modern sequencing-based estimates of $\pi$ across taxa
ranging over only three orders of magnitude ($0.01$--$10\%$,
\cite{Leffler2012-zj}), Lewontin's paradox has persisted unresolved through the
genomics era.

From the beginning, explanations for Lewontin's Paradox have been framed in
terms of the neutralist--selectionist controversy
\citep{Lewontin1974-jb,Kimura1984-ia,Gillespie1991-qa,Gillespie2001-mv}.
The neutralist view is that beneficial alleles are sufficiently rare and
deleterious alleles removed sufficiently quickly, that levels of genetic
diversity are shaped predominantly by genetic drift and mutation
\citep{Kimura1984-ia}. Specifically, \emph{non-selective} processes
decouple the effective population size implied by observed levels of diversity
$\widehat{\pi}$, $\widetilde{N}_e = \nicefrac{\widehat{\pi}}{4\mu}$, from the
census size, $N_c$. By contrast, the selectionist view is that the direct and
indirect effects of linked selection suppress diversity levels across taxa,
specifically because the impact of linked selection is greater in large
populations. Undoubtedly, these opposing views represent a false dichotomy, as
population genomic studies have uncovered both complex demographic histories
that impact diversity within a species (e.g.
\cite{Zhao2013-vd,Palkopoulou2015-bg}) and evidence that selection depresses
genome-wide diversity (e.g.
\cite{Elyashiv2016-vt,Begun1992-ey,Aguade1989-jx,McVicker2009-ax}).

\subsection{Possible Explanations of Lewontin's Paradox}

@@

While chance demographic events like bottlenecks and recent expansions have
long-term impacts on diversity (since mutation-drift equilibrium is reached on
the order of size of the population), characteristics of the breeding structure
such as high variance ($V_w$) or skew in reproductive success continuously
suppress diversity below the levels predicted by the census size
\citep{Wright1938-tv}. For example, in many marine animals, females are
highly fecund, and dispersing larvae face extremely low survivorship, leading
to high variance in reproductive success
\citep{Waples2018-kb,Waples2013-wi,Hedgecock2011-ku,Hauser2008-fd}. Such
``sweepstakes" reproductive systems can lead to remarkably small ratios of
effective population size to census population size (e.g. $\nicefrac{N_e}{N_c}$
can range from $10^{-6}$--$10^{-2}$), since $\nicefrac{N_e}{N} \approx
\nicefrac{1}{V_w}$
\citep{Hedgecock1994-gs,Wright1938-tv,Nunney1993-ef,Nunney1996-wy}, and
require multiple-merger coalescent processes to describe their genealogies
\citep{Eldon2006-ui}. Overall, these reproductive systems diminish the
diversity in many species, but seem unlikely to explain Lewontin's Paradox
broadly across metazoans.

Alternatively, selective processes, and in particular the indirect effects of
selection on linked neutral variation, could explain the observed narrow range
of diversity. The earliest mathematical model of hitchhiking was proffered as a
explanation of Lewontin's Paradox \citep{Maynard_Smith1974-zr}.  Since,
empirical observations have demonstrated that linked selection shapes patterns
of genome-wide diversity, as evidenced by the correlation between recombination
and diversity in a variety of species
\citep{Aguade1989-jx,Begun1992-ey,Cutter2003-tl,Stephan1998-hh,Cai2009-by}.
Theoretic work to explain this pattern has considered diversity under a steady
influx of new beneficial mutations (recurrent hitchhiking;
\cite{Stephan1992-jc,Stephan1995-ry}), and purifying selection against new
deleterious mutations (background selection, BGS;
\cite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}).
Indeed, empirical work indicates background selection diminishes diversity
around genic regions in a variety of species
\citep{McVicker2009-ax,Hernandez2011-gs,Charlesworth1996-px}, and now
efforts have shifted towards teasing apart the effects of positive and negative
selection on genomic diversity \citep{Elyashiv2016-vt}.

An important class of theoretic selection models pertaining to Lewontin's
Paradox are recurrent hitchhiking models that decouple diversity from the
census population size. These models predict diversity when strongly selected
beneficial mutations regularly enter and sweep through the population, trapping
lineages and forcing them to coalesce
\citep{Kaplan1989-sc,Gillespie2000-mh}. In general, decoupling occurs under
these hitchhiking models when the rate of coalescence due to selection is much
greater than the rate of neutral coalescence \citep[e.g.][equation
22]{Coop2012-cd}. Under other linked selection models, the resulting effective
population size is proportional to population size, and thus these models
cannot decouple diversity, all else equal. For example, models of background
selection and polygenic fitness variation predict diversity is proportional to
population size, mediated by the total recombination map length and the
deleterious mutation rate or fitness variation
\citep{Charlesworth1993-gb,Nicolaisen2012-vs,Nordborg1996-nq,Robertson1961-ho,Santiago1995-hx,Santiago1998-bs}.

\subsection{Recent Approaches Towards Solving Lewontin's Paradox}

Recently, \cite{Corbett-Detig2015-gt} used population genomic data to
estimate the reduction in diversity due to background selection and hitchhiking
across 40 species, and showed the impact of selection increases with two
proxies of census population size, species range and with body size. They
argued this is evidence that selection could explain Lewontin's Paradox;
however, in a re-analysis, \cite{Coop2016-gx} demonstrated that the
observed scale of these reductions is insufficient to explain the
orders-of-magnitude shortfall between observed and expected levels of diversity
across species. Other recent work has found that certain life history
characteristics related to parental investment, such as propagule size, are
good predictors diversity in animals \citep{Romiguier2014-bp,Chen2017-nf}.
Nevertheless, while these diversity correlates are important clues, they do not
propose a mechanism by which these traits act to constrain diversity within a
few orders of magnitude.

Here, I revisit Lewontin's Paradox by integrating a variety of data sets and
assessing the predicted reductions in diversity under different selection
models. Prior surveys of genetic diversity either lacked census population size
estimates, used allozyme-based measures of heterozygosity, or included fewer
species. To address these shortcomings, I first estimate census sizes by
combining predictions of population density based on body size with ranges
estimated from geographic occurrence data. Using these estimates, I quantify
the relationship between census size and previously-published genomic diversity
estimates across 172 metazoan taxa within nine phyla, to provide a sense of the
scale of the divergence between $\pi$ and $N_c$ that leads to Lewontin's
Paradox.

Past work looking at the relationship between $\pi$ and $N_c$ has been unable
to fully account for phylogenetic non-independence across taxa
\citep{Felsenstein1985-an}. To address this shortcoming, I use phylogenetic
comparative methods (PCMs) with a synthetic time-calibrated phylogeny to
account for shared phylogenetic history. Moreover, it is disputed whether
considering phylogenetic non-independence is necessary in population genetics,
since coalescent times are much less than divergence times
\citep{Whitney2010-ud,Lynch2011-qv}. Using PCMs, I address this by
estimating the degree of phylogenetic signal in the diversity census size
relationship, and investigating how these traits evolve along the phylogeny.

Finally, I explore whether the predicted reductions of diversity under
background selection and recurrent hitchhiking are sufficiently strong to
resolve Lewontin's Paradox. These predicted reductions in diversity across
species are estimated using strong selection parameters from \emph{Drosophila
melanogaster}, a species known to be strongly affected by linked selection.
Given the effects of linked selection are mediated by recombination map length,
I investigate how recombination map lengths vary with census population size
using data from a previously-published survey \citep{Stapley2017-fs}. I
find map lengths are typically shorter in large--census-size species,
increasing the effects of linked selection in these species, which might
further decouple diversity from census size. Still, I find the combined impact
of these selection models with available parameter estimates falls short in
explaining Lewontin's Paradox, and discuss future avenues through which the
Paradox of Variation could be fully resolved.

@@

I would like to thank Andy Kern and Peter Ralph for helpful discussions and
supporting me during this work, and Graham Coop for inspiration and helpful
feedback during socially distanced nature walks at Yolo Basin. I thank Jessica
Stapley for kindly providing the recombination map length data, and Yaniv
Brandvain, Amy Collins, Doc Edge, Tyler Kent, Chuck Langley, Matt Osmond, Sally
Otto, Jeff Ross-Ibarra, Aaron Stern, Anastasia Teterina, Michael Turelli,
Margot Wood, and my Kern-Ralph labmates for helpful discussions. Sarah
Friedman, Katherine Corn, and Josef Uyeda provided very useful advice about
phylogenetic comparative methods; yet I take full responsibility for any
shortcomings of my analysis. Finally, I am indebted to Guy Sella, Matt Pennell,
and two other anonymous reviewers for helpful feedback. I would like to also
thank UO librarian Dean Walton for helping me track down some rather difficult
to find older papers.  This work was supported by an NIH Grant (1R01GM117241)
awarded to Andrew Kern.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \title{Quantifying Lewontin's Paradox Suggests Natural Selection is Unlikely to Explain the Narrow Range of Diversity Among Species}
\title{Quantifying the Relationship Between Genetic Diversity and Population Size Suggests Natural Selection Cannot Explain Lewontin's Paradox}

@@

 Neutral theory predicts that genetic diversity increases with population size,
 yet observed levels of diversity across metazoans vary only two orders of
 magnitude while population sizes vary over several. This unexpectedly narrow
 range of diversity is known as Lewontin's Paradox of Variation (1974). While
 some have suggested selection constrains diversity, tests of this hypothesis
 seem to fall short. Here, I revisit Lewontin's Paradox to assess whether
 current models of linked selection are capable of reducing diversity to this
 extent. To quantify the discrepancy between pairwise diversity and census
 population sizes across species, I combine previously-published estimates of
 pairwise diversity from 172 metazoan taxa with newly derived estimates of
 census sizes.  Using phylogenetic comparative methods, I show this
 relationship is significant accounting for phylogeny, but with high
 phylogenetic signal and evidence that some lineages experience shifts in the
 evolutionary rate of diversity deep in the past. Additionally, I find a
 negative relationship between recombination map length and census size,
 suggesting abundant species have less recombination and experience greater
 reductions in diversity due to linked selection. However, I show that even
 assuming strong and abundant selection, models of linked selection are
 unlikely to explain the observed relationship between diversity and census
 sizes across species.

\end{abstract}

@@

A longstanding mystery in evolutionary genetics is that the observed levels of
genetic variation across sexual species span an unexpectedly narrow range.
Under neutral theory, the average number of nucleotide differences between
sequences (pairwise diversity, $\pi$) is determined by the balance of new
mutations and their loss by genetic drift
\citep{Kimura1964-ul,Malecot1948-zv,Wright1931-fl}. In particular, expected
pairwise diversity at neutral sites in a panmictic population of $N_c$ diploids
is $\pi \approx 4N_c \mu$, where $\mu$ is the per basepair per generation
mutation rate. Given that metazoan germline mutation rates only differ 10-fold
($10^{-8}$--$10^{-9}$, \cite{Kondrashov2010-fi,Lynch2010-ki}), and census sizes
vary over several orders of magnitude, under neutral theory one would expect
that pairwise diversity also vary over several orders of magnitude.  However,
early allozyme surveys revealed that diversity levels across a wide range of
species varied just an order of magnitude \citep[][p. 208]{Lewontin1974-jb};
this is known as Lewontin's ``Paradox of Variation".  With modern
sequencing-based estimates of $\pi$ across taxa ranging over only three orders
of magnitude ($0.01$--$10\%$, \cite{Leffler2012-zj}), Lewontin's paradox
remains unresolved through the genomics era.

Early on, explanations for Lewontin's Paradox have been framed in terms of the
neutralist--selectionist controversy
\citep{Lewontin1974-jb,Kimura1984-ia,Gillespie1991-qa,Gillespie2001-mv}.  The
neutralist view is that beneficial alleles are sufficiently rare and
deleterious alleles are removed sufficiently quickly, that levels of genetic
diversity are shaped predominantly by genetic drift and mutation
\citep{Kimura1984-ia}. Specifically, \emph{non-selective} processes decouple
the effective population size implied by observed levels of diversity
$\widehat{\pi}$, $\widetilde{N}_e = \nicefrac{\widehat{\pi}}{4\mu}$, from the
census size, $N_c$. By contrast, the selectionist view is that direct selection
and the indirect effects of selection on linked neutral diversity suppress
diversity levels across taxa, specifically because the impact of linked
selection is greater in large populations. Undoubtedly, these opposing views
represent a false dichotomy, as population genomic studies have uncovered
evidence for the substantial impact of both demographic history (e.g.
\cite{Zhao2013-vd,Palkopoulou2015-bg}) and linked selection on genome-wide
diversity (e.g.
\cite{Elyashiv2016-vt,Begun1992-ey,Aguade1989-jx,McVicker2009-ax}).

\subsection{Possible Resolutions of Lewontin's Paradox}

@@

While intermittent demographic events like bottlenecks and recent expansions
have long-term impacts on diversity (since mutation-drift equilibrium is
reached on the order of size of the population), characteristics of the
breeding structure such as high variance ($V_w$) or skew in reproductive
success continuously suppress diversity below the levels predicted by the
census size \citep{Wright1938-tv}. For example, in many marine animals, females
are highly fecund, and dispersing larvae face extremely low survivorship,
leading to high variance in reproductive success
\citep{Waples2018-kb,Waples2013-wi,Hedgecock2011-ku,Hauser2008-fd}. Such
``sweepstakes" reproductive systems can lead to remarkably small ratios of
effective to census population size (e.g. $\nicefrac{N_e}{N_c}$ can range from
$10^{-6}$--$10^{-2}$), since $\nicefrac{N_e}{N} \approx \nicefrac{1}{V_w}$
\citep{Hedgecock1994-gs,Wright1938-tv,Nunney1993-ef,Nunney1996-wy}, and require
multiple-merger coalescent processes to describe their genealogies
\citep{Eldon2006-ui}. Overall, these reproductive systems diminish the
diversity in some species, but seem unlikely to explain Lewontin's Paradox
broadly across metazoans.

Alternatively, selective processes, and in particular the indirect effects of
selection on linked neutral variation, could potentially explain the observed
narrow range of diversity. The earliest mathematical model of hitchhiking was
proffered as a explanation of Lewontin's Paradox \citep{Maynard_Smith1974-zr}.
Since, linked selection has been shown to impact diversity levels in a variety
of species, as evidenced by the correlation between recombination and diversity
\citep{Aguade1989-jx,Begun1992-ey,Cutter2003-tl,Stephan1998-hh,Cai2009-by}.
Theoretic work to explain this pattern has considered the impact of a steady
influx of beneficial mutations (recurrent hitchhiking;
\cite{Stephan1992-jc,Stephan1995-ry}), and purifying selection against
deleterious mutations (background selection, BGS;
\cite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}).
Indeed, empirical work indicates background selection diminishes diversity
around genic regions in a variety of species
\citep{McVicker2009-ax,Hernandez2011-gs,Charlesworth1996-px}, and now efforts
have shifted towards teasing apart the effects of positive and negative
selection on genomic diversity \citep{Elyashiv2016-vt}.

A class of models that are of particular interest in the context of Lewontin's
Paradox are recurrent hitchhiking models that decouple diversity from the
census population size. These models predict diversity levels when strongly
selected beneficial mutations regularly enter and sweep through the population,
trapping lineages and forcing them to coalesce
\citep{Kaplan1989-sc,Gillespie2000-mh}. In general, decoupling occurs under
these hitchhiking models when the rate of coalescence due to selection is much
greater than the rate of neutral coalescence \citep[e.g.][equation
22]{Coop2012-cd}. In contrast, under other linked selection models, the
resulting effective population size is proportional to population size; these
models cannot decouple diversity, all else equal. For example, models of
background selection and polygenic fitness variation predict diversity is
proportional to population size, mediated by the total recombination map length
and the deleterious mutation rate or fitness variation
\citep{Charlesworth1993-gb,Nicolaisen2012-vs,Nordborg1996-nq,Robertson1961-ho,Santiago1995-hx,Santiago1998-bs}.

\subsection{Recent Approaches Towards Resolving Lewontin's Paradox}

Recently, \cite{Corbett-Detig2015-gt} used population genomic data to estimate
the reduction in diversity due to background selection and hitchhiking across
40 species, and showed that the impact of selection increases with two proxies
of census population size, species range and with body size. Based on this
evidence, they argued that selection could explain Lewontin's Paradox; however,
in a re-analysis, \cite{Coop2016-gx} demonstrated that the observed magnitude
of these reductions is insufficient to explain the orders-of-magnitude
shortfall between observed and expected levels of diversity across species.
Other recent work has found that life history characteristics related to
parental investment, such as propagule size, are good predictors diversity in
animals \citep{Romiguier2014-bp,Chen2017-nf}.  Nevertheless, while these
diversity correlates are important clues, they do not propose a mechanism by
which these traits act to constrain diversity within a few orders of magnitude.

Here, I revisit Lewontin's Paradox by integrating several data sets in order to
compare the observed reductions between diversity and census size with the
predicted relationship under different selection models. Prior surveys of
genetic diversity either lacked census population size estimates, used
allozyme-based measures of heterozygosity, or included fewer species. To
address these shortcomings, I first estimate census sizes by combining
predictions of population density based on body size with ranges estimated from
geographic occurrence data. Using these estimates, I quantify the relationship
between census size and previously-published genomic diversity estimates across
172 metazoan taxa within nine phyla, thus characterizing the relationship
between $\pi$ and $N_c$ that underlies Lewontin's Paradox.

Past work looking at the relationship between $\pi$ and $N_c$ has been unable
to fully account for phylogenetic non-independence across taxa
\citep{Felsenstein1985-an}. To address this, I use phylogenetic comparative
methods (PCMs) with a synthetic time-calibrated phylogeny to account for shared
phylogenetic history. Moreover, it is disputed whether considering phylogenetic
non-independence is necessary in population genetics, given that coalescent
times within species are much less than divergence times
\citep{Whitney2010-ud,Lynch2011-qv}. Using PCMs, I address this by estimating
the degree of phylogenetic signal in the diversity census size relationship,
and investigating how these traits evolve along the phylogeny.

Finally, I explore whether the predicted reductions of diversity under
background selection and recurrent hitchhiking are sufficiently strong to
resolve Lewontin's Paradox. I do so using selection parameters from
\emph{Drosophila melanogaster}, a species known to be strongly affected by
linked selection. Given the effects of linked selection are mediated by
recombination map length, I also investigate how recombination map lengths vary
with census population size using data from a previously-published survey
\citep{Stapley2017-fs}. I find map lengths are typically shorter in
large--census-size species, increasing the effects of linked selection in these
species, which could further decouple diversity from census size. Still, I find
the combined impact of these modes of linked selection fall short in explaining
Lewontin's Paradox, and discuss future avenues through which the Paradox of
Variation could be fully resolved.

@@

I would like to thank Andy Kern and Peter Ralph for helpful discussions and
supporting me during this work, and Graham Coop for inspiration and helpful
feedback during socially distanced nature walks at Yolo Basin. I thank Jessica
Stapley for kindly providing the recombination map length data, and Yaniv
Brandvain, Amy Collins, Doc Edge, Tyler Kent, Chuck Langley, Matt Osmond, Sally
Otto, Molly Przeworski, Jeff Ross-Ibarra, Aaron Stern, Anastasia Teterina,
Michael Turelli, Margot Wood, and my Kern-Ralph labmates for helpful
discussions. Sarah Friedman, Katherine Corn, and Josef Uyeda provided very
useful advice about phylogenetic comparative methods; yet I take full
responsibility for any shortcomings of my analysis. Finally, I am indebted to
Guy Sella, Matt Pennell, and two other anonymous reviewers for helpful
feedback. I would like to also thank UO librarian Dean Walton for helping me
track down some rather difficult to find older papers.  This work was supported
by an NIH Grant (1R01GM117241) awarded to Andrew Kern.
